---
title: "The Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Jonathan Bouchet"
date: "November 21, 2015"
output: pdf_document
---
##*<span style="color:red">Instruction<span>*

1. Load the ToothGrowth data and perform some basic exploratory data analyses 

2. Provide a basic summary of the data.

3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)

4. State your conclusions and the assumptions needed for your conclusions. 

##*<span style="color:red">Preambule<span>*
```{r}
library(datasets)
data(ToothGrowth)
```
A brief explanation of the data is given in the help :
```{r eval=FALSE}  
help(ToothGroth)
```

An overview of the data shows that it consists of a dataframe with 60 observations on 3 variables :

```{r eval=FALSE}
>str(ToothGrowth)
'data.frame':	60 obs. of  3 variables:
 $ len : num  4.2 11.5 7.3 5.8 6.4 10 11.2 11.2 5.2 7 ...
 $ supp: Factor w/ 2 levels "OJ","VC": 2 2 2 2 2 2 2 2 2 2 ...
 $ dose: num  0.5 0.5 0.5 0.5 0.5 0.5 0.5 0.5 0.5 0.5 ...
>summary(ToothGrowth)
      len        supp         dose 
 Min.   : 4.20   OJ:30   Min.   :0.500  
 1st Qu.:13.07   VC:30   1st Qu.:0.500  
 Median :19.25           Median :1.000  
 Mean   :18.81           Mean   :1.167  
 3rd Qu.:25.27           3rd Qu.:2.000  
 Max.   :33.90           Max.   :2.000 
 ```
We can summarize the data with the following plot :
```{r}
boxplot(len~supp*dose, data=ToothGrowth,col=(c("plum1","lightblue")),main="Tooth Growth", xlab="Supplement and Dose",ylab="length")
```

###*<span style="color:blue">Comments :<span>*
1. We see that the length increase in both delivery method ($\textbf{OJ}$: orange juice or $\textbf{VC}$:vitamin C ) when the dose increase

2. $\textbf{OJ}$ supplement is always above $\textbf{VC}$ for a given dose, meaning that it is more effective than $\textbf{VC}$. It seems to be especially true for the low dose level.


To make the manipulation of the data :
```{r}
attach(ToothGrowth)
```

##*<span style="color:red">Analysis<span>*

We split the original dataset into 6 columns (3 doses, 2 supplements) before doing the confidence intervals with t-tests, and we rename them.

```{r}
DoseGroups<-as.data.frame(split(len,list(supp,dose)))
names(DoseGroups) <- c('OJ_d05','VC_d05','OJ_d1','VC_d1','OJ_d2','VC_d2')
```

```{r eval = FALSE}
> str(DoseGroups)
'data.frame':	10 obs. of  6 variables:
 $ OJ_d05: num  15.2 21.5 17.6 9.7 14.5 10 8.2 9.4 16.5 9.7
 $ VC_d05: num  4.2 11.5 7.3 5.8 6.4 10 11.2 11.2 5.2 7
 $ OJ_d1 : num  19.7 23.3 23.6 26.4 20 25.2 25.8 21.2 14.5 27.3
 $ VC_d1 : num  16.5 16.5 15.2 17.3 22.5 17.3 13.6 14.5 18.8 15.5
 $ OJ_d2 : num  25.5 26.4 22.4 24.5 24.8 30.9 26.4 27.3 29.4 23
 $ VC_d2 : num  23.6 18.5 33.9 25.5 26.4 32.5 26.7 21.5 23.3 29.5
```

Now our data consists of 10 observations (pigs) for each of the configurations (6 = 3 doses $\times$ 2 supplements).  
As the number of observations are not high enough with perform t-tests with assumption that these are independent groups (i.e., not paired).

As the number of observations are the same for each group, it doesn’t make a difference whether the variances are equal for different groups or not. We do all the analysis assuming unequal variance.

The null hypothesis tested ($H_{0}$) is : there is no difference in the mean for each supplement and therefore $\textbf{OJ}$ and $\textbf{VC}$ have the same effect on tooth growth.

We display the results of the lower ("low") and upper ("up") value of the Confidence Interval, as well as the p-value for each dose.
<!--
###*<span style="color:blue">dose = 0.5 mg<span>*
```{r}
t.test(DoseGroups['OJ_d05'] - DoseGroups['VC_d05'])$conf
t.test(DoseGroups['OJ_d05'] - DoseGroups['VC_d05'])$p.value
```

###*<span style="color:blue">dose = 1.0 mg<span>*
```{r}
t.test(DoseGroups['OJ_d1'] - DoseGroups['VC_d1'],paired = FALSE,var.equal = FALSE)$conf
t.test(DoseGroups['OJ_d1'] - DoseGroups['VC_d1'],paired = FALSE,var.equal = FALSE)$p.value
```

###*<span style="color:blue">dose = 2.0 mg<span>*
```{r}
t.test(DoseGroups['OJ_d2'] - DoseGroups['VC_d2'],paired = FALSE,var.equal = FALSE)$conf
t.test(DoseGroups['OJ_d2'] - DoseGroups['VC_d2'],paired = FALSE,var.equal = FALSE)$p.value
```
-->

```{r}
i<-1;
counter<-1;
M<-matrix(ncol=3,nrow=3);
dimnames(M) = list(c("dose=0.5 mg","dose=1.0 mg","dose=2.0 mg"),c("low","up","p-value"));
while ( i < 6) {
  M[counter,1]<-t.test(DoseGroups[i] - DoseGroups[i+1],paired = F,var.equal = F)$conf.int[1]; 
  M[counter,2]<-t.test(DoseGroups[i] - DoseGroups[i+1],paired = F,var.equal = F)$conf.int[2]; 
  M[counter,3]<-t.test(DoseGroups[i] - DoseGroups[i+1],paired = F,var.equal = F)$p.value;
  i<-i+2;
  counter<-counter+1;}
print(M)
```

We see that :

1. the confidence levels for dose = 0.5 mg and dose = 1.0 mg do not contain zero, and the p-value is lower than 0.05. In these 2 cases we can reject the $H_{0}$. Thus the difference in the mean we had seen with the plot between the 2 supplements is real.

2. For the last dose, the Confidence interval does contains zero with a p-value>.95, thus we cannot reject the null hypothesis, thus we cannot conclude on the efffectiveness of the impact between the 2 supplements for this dose.
<!--
```{r}
i<-1;
counter<-1;
res<-vector(length=3);
names(res)<-c("p-value dose=0.5 mg","p-value dose=1.0 mg","p=value dose=2.0 mg");
while ( i < 6) {res[counter]<-t.test(DoseGroups[i] - DoseGroups[i+1],paired = F,var.equal = F)$p.value;i<-i+2;counter<-counter+1;}
res;
```
-->

###*<span style="color:blue">notes<span>* 
It is also important to notice than when comparing data from different doses and/or different, the confidence interval are always negative. For example :
```{r}
t.test(DoseGroups['OJ_d05'] - DoseGroups['VC_d1'],paired = F,var.equal = F)$conf
t.test(DoseGroups['OJ_d05'] - DoseGroups['VC_d2'],paired = F,var.equal = F)$conf
```
